---
title: "NFL Passing Analysis"
author: "Tucker Morgan"
date: "1/9/2022"
output: html_document
---

```{css, echo = FALSE}
pre {
  max-height: 150px;
  overflow-y: auto;
}

pre[class] {
  max-height: 150px;
}

.scroll-100 {
  max-height: 150px;
  overflow-y: auto;
  background-color: inherit;
}
```

```{r setup, include=FALSE}
library(tidyverse)
library(rvest)
library(RColorBrewer)
library(viridis)
```

## Introduction

### Purpose

Teams in the National Football League (NFL) are racking up more passing yards than ever, so how can current quarterback (QB) passing stats be evaluated against historical data? This analysis will aim to answer that question and provide insight into how passing stats and QB performances have changed over the past 25 years.

I originally set out to analyze these data during my first semester at Columbia as I learned more R. Well, turns out we learned even more helpful material later in the semester, so I'm planning to run a re-vamped or "Reloaded" version now. I'm hoping to streamline some of my code from last time (for example, using `for` loops instead of copying code).

All of the data used in this analysis was obtained from https://www.pro-football-reference.com/.

### Motivation

A common theme over the past few years in NFL-commentary circles has been how to compare current NFL QB performances (particularly seasonal stat totals) to an historic benchmark. The game has changed in many ways from playstyle philosophies to major rule changes, so it's a fair question: do higher stats (like yards or touchdowns) truly indicate better performing QBs or is it easier / more common to rack up stats in "today's NFL"? A particular podcast highlighting this in my mind has been [The Ryen Russillo Podcast](https://www.theringer.com/ryen-russillo-podcast) from The Ringer. A common refrain has been something along the lines of "well, 4,000 yards ain't what it used to be."

I will analyze the distribution of passing yards over the years and see if classic benchmarks like 4,000 yards passing in a single season are as meaningful now as they were 25 years ago. I will also propose a simple measure to assess current performances against those of yesteryear.

## Data Import and Cleaning

I want to first set up a function to load in and clean each yearly dataset.

```{r url test}
url <- "https://www.pro-football-reference.com/years/2021/passing.htm"
nfl_pass_html <- read_html(url)

nfl_pass_table <-
  nfl_pass_html %>%
  html_table() %>%
  first() %>% 
  janitor::clean_names() %>% 
  mutate(year = "2021")

rm(nfl_pass_html)
# Here I was able to successfully load in and extract the table from pro-football-reference;
# in my function, I will want to do this for each year and add a column for year. I think this can be incorporated into the function
```

```{r import function}
read_nfl_passing = function(url){
  
  html = read_html(url)
  
  pass_table = html %>% 
    html_table() %>% 
    first() %>% 
    janitor::clean_names()
  
  output = pass_table
  
}
```

```{r using map and dataframe}
nfl_passing <- 
  tibble(
    season = 1995:2021,
    urls = str_c("https://www.pro-football-reference.com/years/", season, "/passing.htm")) %>% 
  mutate(stats = map(urls, read_nfl_passing)) %>% 
  unnest(cols = c(stats))
```

Nice! I was able to pull in several seasons of data from Pro Football Reference. It looks like there are a few columns I can remove to streamline the dataset from here, and I think the function pulled in the column headers for the additional pages of each year. I'll go ahead and filter out the additional headers and unneeded columns. I will also need to convert the columns to numbers and factors accordingly, since all were considered characters because of the intermediary headers.

```{r data cleaning}
nfl_passing_clean <- 
  nfl_passing %>% 
  filter(rk != "Rk") %>% 
  select(-urls, -pos, -td_percent, -int_percent, -x1d, -lng, -y_c, -y_g, -rate, -sk, -yds_2, -sk_percent, -ny_a, -any_a, -x4qc, -gwd, -qbr) %>% 
  rename(rank = rk, team = tm, games = g, starts = gs, record = q_brec, completions = cmp, attempts = att, yards = yds) %>% 
  mutate(rank = as.numeric(rank),
         team = as.factor(team),
         age = as.numeric(age),
         games = as.numeric(games),
         starts = as.numeric(starts),
         completions = as.numeric(completions),
         attempts = as.numeric(attempts),
         cmp_percent = as.numeric(cmp_percent),
         yards = as.numeric(yards),
         td = as.numeric(td),
         int = as.numeric(int))
```

This data set includes every player to attempt a pass in each NFL season, however we are primarily interested in "the main guys" over the years. We'd like our data set to be indicative of QBs who are playing consistently from year to year - career QBs. First, let's look at the distribution of yards and attempts.

```{r yards distribution}
nfl_passing_clean %>% 
  ggplot(aes(x = yards)) +
  geom_density()

nfl_passing_clean %>% 
  ggplot(aes(x = attempts)) +
  geom_density()

nfl_passing_clean %>% 
  ggplot(aes(x = rank)) + geom_density()
```

As expected, there are a large number of players with fewer than 1,000 or even 500 yards and a large number of players with fewer than 100 passing attempts. Since we're interested in how performances change over time, we may want to be careful about our cut-off here. Setting a minimum yards or attempts cut-off could include more quarterbacks in future years compared to earlier years. In each season, Pro Football Reference has QBs ranked from 1 to n with n varying each year depending on how many players attempt a pass. I think setting a cut-off for `rank` might be the most consistent cut-off from year-to-year.
